---
title: "Visualization Design Canvas"
author: "Magdalena A. Tkacz"
subtitle: "Template"
output:
  pdf_document:
    toc: true
    toc_depth: 3
  html_document:
    df_print: paged
    theme: spacelab
    highlight: zenburn
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    code_fold: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "figures/"
)
```

# Example Project

# Gapminder — From Data to Visual Language

> *Good visualizations are designed, not generated.*

---

# DN-001 — Dataset Card

## Dataset

**Title**

Gapminder

**Source**

Gapminder Foundation

https://www.gapminder.org/

**Original author**

Gapminder Foundation

---

## Description

The Gapminder dataset contains demographic, economic and health indicators collected for countries around the world over multiple years.

Each observation represents one country in one year.

The dataset combines several variables that allow exploration of relationships between wealth, health and population over time.

---

## Observations

One row = one country in one year.

---

## Variables

| Variable | Type | Description |
|----------|------|-------------|
| country | categorical | Country name |
| continent | categorical | Continent |
| year | temporal | Observation year |
| lifeExp | continuous | Life expectancy |
| pop | continuous | Population |
| gdpPercap | continuous | GDP per capita |

---

## Data quality

### Missing values

Very few.

### Outliers

Countries with extremely large populations.

Countries with very high GDP per capita.

### Potential bias

Countries differ greatly in population size.

Economic indicators are estimates.

### Initial observations

Health and wealth seem related.

Population varies by several orders of magnitude.

Time appears to play an important role.

---

# DN-002 — Audience & Goal Card

## Audience

Students interested in data visualization.

Basic statistical knowledge.

Little experience with visualization design.

---

## Context

Presentation during a workshop.

---

## Main objective

Explain the relationship between economic development and health.

---

## Key message

Countries become healthier as wealth increases, but development is not uniform.

---

## Main question

How do health, wealth and population change over time?

---

## Success criteria

The audience should immediately recognize:

- relationship between GDP and life expectancy
- differences between continents
- population size
- temporal evolution

---

# DN-003 — Data Understanding Card

## Variables of interest

| Variable | Type | Importance |
|----------|------|------------|
| GDP per capita | continuous | High |
| Life expectancy | continuous | High |
| Population | continuous | High |
| Continent | categorical | High |
| Year | temporal | High |


## Relationships

GDP vs. Life expectancy

Population vs/ Country importance

Continent vs. Regional differences

Year vs. Development



## Interesting questions

How does wealth influence life expectancy?

Do continents evolve differently?

Which countries improve the fastest?

What happens through time?



## Hypothesis

Countries generally become healthier and wealthier over time.



# DN-004 — Visualization Mapping Card

## Data → Visual Encoding

| Data | Encoding | Why |
|------|----------|-----|
| GDP per capita | X position | Accurate quantitative comparison |
| Life expectancy | Y position | Accurate quantitative comparison |
| Population | Bubble size | Magnitude |
| Continent | Color | Category recognition |
| Year | Animation | Temporal evolution |



## Visual hierarchy

### First

Overall relationship between wealth and health.

### Second

Large countries.

### Third

Continental patterns.



## Accessibility

✔ Colorblind-friendly palette

✔ High contrast

✔ Minimal clutter

✔ Simple labels


# DN-005 — Representation Card

## Candidate representations

### Option A

Scatter plot

Advantages

- relationship visible
- simple
- accurate

Limitations

- no population
- no time



### Option B

Bubble chart

Advantages

- adds population

Limitations

- bubble area harder to compare



### Option C

Animated bubble chart

Advantages

- shows temporal evolution
- engaging
- reveals trajectories

Limitations

- harder to compare specific years



## Decision

Animated Bubble Chart

Reason:

It communicates five variables simultaneously while maintaining readability.



# DN-006 — Implementation Card

## Tool

R

ggplot2

gganimate



## Why?

Grammar of Graphics provides a declarative way to map data onto visual attributes.

Animation naturally represents temporal change.



## Workflow

1. Import data

2. Explore variables

3. Design mapping

4. Prototype

5. Refine

6. Review



# DN-007 — Review Card

## Message

✔ Clear

✔ Focused

✔ Honest



## Audience

✔ Appropriate

✔ Easy to understand



## Design

✔ Effective color

✔ Good visual hierarchy

✔ Appropriate labels

✔ Limited chartjunk



## Reflection

Strengths

- multiple variables
- intuitive mapping
- memorable

Weaknesses

- animation prevents precise comparison
- bubble area is less accurate than position

Possible improvements

- small multiples
- interactive controls
- annotations


# Final Design Summary

## Goal

Explain global development.

## Audience

Students learning visualization.

## Question

How are wealth, health and population related over time?

## Representation

Animated bubble chart.

## Mapping

GDP → X

Life expectancy → Y

Population → Size

Continent → Color

Year → Animation

## Implementation

ggplot2 + gganimate


> "The visualization was not chosen because it looks attractive.
>
> It was chosen because every visual attribute supports the communication goal."

### Implementation details

#### AI prompt i implementation code.

##### DN-008 — Implementation Prompt

*Exemplary prompt for AI-assisted implementation:*

> Create an R implementation of an animated Gapminder bubble chart.
>
> Use the `gapminder` dataset.
>
> The visualization should show:
>
> - GDP per capita on the X axis,
> - life expectancy on the Y axis,
> - population as bubble size,
> - continent as color,
> - year as animation frame.
>
> Use `ggplot2` and `gganimate`.
>
> Use a log scale for GDP per capita.
>
> Make the chart readable, clean and suitable for a presentation.
>
> Add clear title, subtitle, axis labels and legend labels.
>
> Export the animation as a GIF file.

Remarks:

- A prompt should describe design decisions, not code. 
- The code is simply the implementation of those decisions in R.

##### DN-009 — R Implementation

```{r eval=FALSE, include=FALSE}
# ============================================================
# Gapminder Animated Bubble Chart
# Design-driven implementation
# ============================================================

# Install packages if needed:
# install.packages(c("gapminder", "ggplot2", "gganimate", "gifski"))

library(gapminder)
library(ggplot2)
library(gganimate)
library(gifski)

# ------------------------------------------------------------
# 1. Data
# ------------------------------------------------------------

data <- gapminder

# ------------------------------------------------------------
# 2. Static mapping
# ------------------------------------------------------------

p <- ggplot(
  data,
  aes(
    x = gdpPercap,
    y = lifeExp,
    size = pop,
    color = continent
  )
) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  scale_size(
    range = c(2, 18),
    name = "Population"
  ) +
  labs(
    title = "Health and Wealth of Nations",
    subtitle = "Year: {frame_time}",
    x = "GDP per capita (log scale)",
    y = "Life expectancy",
    color = "Continent",
    caption = "Data source: Gapminder"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# ------------------------------------------------------------
# 3. Animation
# ------------------------------------------------------------

animated_plot <- p +
  transition_time(year) +
  ease_aes("linear")

# ------------------------------------------------------------
# 4. Render animation
# ------------------------------------------------------------

animate(
  animated_plot,
  width = 900,
  height = 600,
  fps = 8,
  duration = 12,
  renderer = gifski_renderer("gapminder_animation.gif")
)
```


##### N-010 — Static Alternative

```{r eval=FALSE, include=FALSE}
library(gapminder)
library(ggplot2)

data_2007 <- subset(gapminder, year == 2007)

ggplot(
  data_2007,
  aes(
    x = gdpPercap,
    y = lifeExp,
    size = pop,
    color = continent
  )
) +
  geom_point(alpha = 0.75) +
  scale_x_log10() +
  scale_size(
    range = c(2, 18),
    name = "Population"
  ) +
  labs(
    title = "Health and Wealth of Nations in 2007",
    subtitle = "GDP per capita, life expectancy and population",
    x = "GDP per capita (log scale)",
    y = "Life expectancy",
    color = "Continent",
    caption = "Data source: Gapminder"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

# Final Thoughts

Today, the most valuable skill is no longer writing visualization code. Instead it is making and communicating sound design decisions. The code is merely an implementation; the design is where visualization begins.


The key competency of the future is not mastering programming syntax or visualization tools, but making sound design decisions and expressing visualization requirements with clarity.

